The demographic costs of sexually antagonistic selection in partially selfing populations

### 2021-12-30 doc/SA-Hermaphrodites-wDemography-R1.tex
Updated table 1 and added one sentence about maturation rate

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[htbp]
\centering
\caption{\bf Definition of terms.}
\label{tab:Terms}
\begin{tabular}{ l p{0.7\linewidth} r }
 \toprule
Symbol & Definition & Dimension \\
\hline
$g$      & Number of genotypes ($3$; $AA$, $Aa$, and $aa$) & \\
$\omega$ & Number of stages ($2$; juvenile and adult) & \\
$\tilde{\mbf{n}}$ & Joint stage $\times$ genotype vector & $2 \omega g \times 1$ \\
$\tilde{\mbf{p}}$ & Joint stage $\times$ genotype frequency vector & $2 \omega g \times 1$ \\
$\mbf{U}^{S}_{i}$,\, $\mbf{U}^{X}_{i}$ & Genotype-specific transition and survival matrices & $\omega \times \omega$ \\
$\mcal{U}^{S},\, \mcal{U}^{X}$ & Block diagonal selfed/outcrossed survival matrices & $\omega g \times \omega$ g\\
$\mbf{F}_{i}$,\, $\mbf{F}^{\prime}_{i}$ & Genotype-specific fertility matrices & $\omega \times \omega$ \\
$\mcal{F}^{S},\, \mcal{F}^{X}$ & Block diagonal selfed/outcrossed fertility matrix & $\omega g \times \omega$g \\
% $\mbf{K}$   & Vec-permutation matrix & $g \times g$ \\
% $\mbf{I}$   & Identity matrix & given in subscript \\
$q^{\prime}_{A}$,\, $q^{\prime}_{a}$ & Allele frequencies in male gamete pool & $1$ \\
$\tilde{\mbf{A}}(\tilde{\mbf{p}})$ & Population projection matrix & $2 \omega g \times 2\omega g$ \\

$s_f,\,s_m$ & Selection coefficients through female and male reproductive function & \\
$h_f,\,h_m$ & Dominance coefficients through female and male reproductive function & \\
$\sigma_j,\, \sigma_a$ & Survival rates for juvenile and adult stages & \\
$\gamma$ & Transition rate from juvenile to adult stages \\
$C$ & The population rate of self-fertilization & \\
$\delta$ &\textcolor{blue}{ Inbreeding depression effect on ovule viability}&\\
$ \delta_j$ & \textcolor{blue}{ Inbreeding depression effect on juvenile survival}& \\
$\delta_a$ & \textcolor{blue}{ Inbreeding depression effect on adult survival}& \\
$\delta_{\gamma}$ &\textcolor{blue}{Inbreeding depression effect on maturation rate} & \\
$f_i,\,f^{\prime}_i$ & Adult fertilities through female and male sex functions respectively & \\

@@

To analyze the demographic effects of inbreeding depression, we make two additional simplifying assumptions: First, to keep our analyses tractable we explore the effects of individual inbreeding depression terms in isolation. That is, we assume that only one of the $\delta$ and $\delta_i$ terms (where $i \in \{j,a,\gamma\}$) can be non-zero at a time. Second, we assume that if inbreeding depression is caused primarily by recessive deleterious mutations (as suggested by empirical data), it should covary negatively with the population selfing rate due to purging, provided the population selfing rate has been relatively constant in recent evolutionary time (\citealt{Charlesworth2009}; though we note that other processes could give rise to this pattern, e.g., \citealt{CrnokrakBarrett2002, Charlesworth2009,HedrickGarcia-Dorado2016}). Following \citet{Olito2019}, we incorporate such negative covariance by constraining the inbreeding depression terms in the model ($\delta$ and $\delta_i$, where $i \in \{j,a,\gamma\}$) to follow a simple declining function of the selfing rate: \textcolor{blue}{$\delta = \delta^{\ast} (1 - b (1 - a(1 - C)/(C + a(1 - C))))$}, where $\delta^{\ast}$ is the hypothetical severity of inbreeding depression for a completely outcrossing population, $b$ is a shape parameter determining how far $\delta$ will decline under complete selfing (when $C = 1$), \textcolor{blue}{and $a$ is a shape parameter} controlling the curvature of the overall function for $\delta$ (see Appendix E in \citealt{Olito2019} for additional details). We set $\delta^{\ast} = 0.8$, $b = 0.5$, and $a = 0.2$ for all analyses involving inbreeding depression, values chosen to be consistent with empirical estimates of inbreeding depression (e.g., fig.~2 in \citealt{HusbandSchemske1996}).

@@

Unlike previous population genetic models, which assume constant population sizes \citep{JordanConnallon2014,Olito2017}, mortality caused by inbreeding depression can strongly influence population persistence in our evolutionary demographic model. Populations with high fertility rates can maintain positive population growth rates despite this additional mortality. This causes a greater proportion of SA polymorphic parameter space to be demographically viable, with the demographic model predictions converging on those from population genetic models in high-fertility populations (fig.~\ref{fig:deltaPolySpace}). For populations with lower fertility rates, the proportion of demographically viable polymorphic parameter space \textcolor{blue}{is very similar to} the population genetic model predictions \textcolor{blue}{for low selfing rates}. \textcolor{blue}{Note, however, that in our model, mortality due to late-acting inbreeding depression also reduces the population effective selfing rate, and therefore shifts the invasion boundaries for the SA alleles to more closely resemble an outcrossing population. In some cases, this can lead to larger regions of viable polymorphic parameter space than predicted by the population genetic models, which can only include early-acting inbreeding depression (e.g., see cross symbols in fig.~\ref{fig:deltaPolySpace})}. However, as the selfing rate increases, demographic viability eventually crashes when the population can no longer sustain the concomitant increase in mortality due to inbreeding depression (fig.~\ref{fig:deltaPolySpace}).

@@

The point in the life-cycle where inbreeding depression manifests can influence the threshold selfing rate at which demographically viable polymorphic parameter space crashes. Our results indicate that population viability was most sensitive to inbreeding depression affecting juvenile survival rates ($\delta_j$; fig.~\ref{fig:deltaPolySpace}, dark blue points), while early-acting inbreeding depression ($\delta$, ovule abortion shortly after fertilization) had a similar effect on population viability as late-acting inbreeding depression affecting adult survial ($\delta_a$) and juvenile-to-adult transition rates ($\delta_{\gamma}$). Inbreeding effects on juvenile survival had the strongest effect on population viability because on average individuals will spend multiple time steps in the juvenile stage before they mature. At each time, juveniles have a probability $\gamma$ to mature and a probability $\sigma_j(1-\delta_j)$ to survive. Inbreeding depression at the juvenile stage therefore makes it harder to survive long enough to mature. Although early-acting inbreeding depression ($\delta$) actually manifests earlier in the life-cycle than juvenile survival, it acts only once by influencing the total number of self-fertilized zygotes that become juveniles. Note, however, that the relative strength of inbreeding depression at different stages of the life history will depend on the interaction between the various inbreeding depression parameters, which we have precluded from our analyses.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[htbp]
\centering
\caption{\bf Mathematical notation and parameter values used in this article.}
\label{tab:Terms}
\begin{tabular}{ l p{0.7\linewidth} r }
 \toprule
\multicolumn{3}{l}{Definition of symbols used in the general model description} \\

Symbol & Definition & Dimension \\
\hline
$g$      & Number of genotypes ($3$; $AA$, $Aa$, and $aa$) & \\
$\omega$ & Number of stages ($2$; juvenile and adult) & \\
$\tilde{\mbf{n}}$ & Joint stage $\times$ genotype vector & $2 \omega g \times 1$ \\
$\tilde{\mbf{p}}$ & Joint stage $\times$ genotype frequency vector & $2 \omega g \times 1$ \\
$\mbf{U}^{S}_{i}$,\, $\mbf{U}^{X}_{i}$ & Genotype-specific transition and survival matrices & $\omega \times \omega$ \\
$\mcal{U}^{S},\, \mcal{U}^{X}$ & Block diagonal selfed/outcrossed survival matrices & $\omega g \times \omega$ g\\
$\mbf{F}_{i}$,\, $\mbf{F}^{\prime}_{i}$ & Genotype-specific fertility matrices & $\omega \times \omega$ \\
$\mcal{F}^{S},\, \mcal{F}^{X}$ & Block diagonal selfed/outcrossed fertility matrix & $\omega g \times \omega$g \\
% $\mbf{K}$   & Vec-permutation matrix & $g \times g$ \\
% $\mbf{I}$   & Identity matrix & given in subscript \\
$q^{\prime}_{A}$,\, $q^{\prime}_{a}$ & Allele frequencies in male gamete pool & $1$ \\
$\tilde{\mbf{A}}(\tilde{\mbf{p}})$ & Population projection matrix & $2 \omega g \times 2\omega g$ \\
\hline
\multicolumn{3}{l}{Parameters used in the two stage example} \\
Symbol & Definition & Value/Range\\
\hline
$s_f,\,s_m$ & Selection coefficients through female and male reproductive function & 0-0.15\\
$h_f,\,h_m$ & Dominance coefficients through female and male reproductive function & $\frac{1}{2}$ or $\frac{1}{4}$\\
$\sigma_j,\, \sigma_a$ & Survival rates for juvenile and adult stages & 0.6\\
$\gamma$ & Transition rate from juvenile to adult stages & 0.05\\
$C$ & The population rate of self-fertilization &0-1 \\
$\delta$ &\textcolor{blue}{ Inbreeding depression effect on ovule viability}&0-0.8\\
$ \delta_j$ & \textcolor{blue}{ Inbreeding depression effect on juvenile survival}& 0-0.8\\
$\delta_a$ & \textcolor{blue}{ Inbreeding depression effect on adult survival}& 0-0.8\\
$\delta_{\gamma}$ &\textcolor{blue}{Inbreeding depression effect on maturation rate} & 0-0.8\\
$f_i,\,f^{\prime}_i$ & Adult fertilities through female and male sex functions respectively & 5.8-8.5 \\

@@

To analyze the demographic effects of inbreeding depression, we make two additional simplifying assumptions: First, to keep our analyses tractable we explore the effects of individual inbreeding depression terms in isolation. That is, we assume that only one of the $\delta$ and $\delta_i$ terms (where $i \in \{j,a,\gamma\}$) can be non-zero at a time. Second, we assume that if inbreeding depression is caused primarily by recessive deleterious mutations (as suggested by empirical data), it should covary negatively with the population selfing rate due to purging, provided the population selfing rate has been relatively constant in recent evolutionary time (\citealt{Charlesworth2009}; though we note that other processes could give rise to this pattern, e.g., \citealt{CrnokrakBarrett2002, Charlesworth2009,HedrickGarcia-Dorado2016}). Following \citet{Olito2019}, we incorporate such negative covariance by constraining the inbreeding depression terms in the model ($\delta$ and $\delta_i$, where $i \in \{j,a,\gamma\}$) to follow a simple declining function of the selfing rate: \textcolor{blue}{
\begin{equation}
  \delta = \delta^{\ast} \left(1 - b \left(1 - \frac{a(1 - C)}{(C + a(1 - C))}\right)\right),
  \label{eq:Canddelta}
\end{equation}}
where $\delta^{\ast}$ is the hypothetical severity of inbreeding depression for a completely outcrossing population, $b$ is a shape parameter determining how far $\delta$ will decline under complete selfing (when $C = 1$), \textcolor{blue}{and $a$ is a shape parameter} controlling the curvature of the overall function for $\delta$ (see Appendix E in \citealt{Olito2019} for additional details). We set $\delta^{\ast} = 0.8$, $b = 0.5$, and $a = 0.2$ for all analyses involving inbreeding depression, values chosen to be consistent with empirical estimates of inbreeding depression (e.g., fig.~2 in \citealt{HusbandSchemske1996}).

@@

Unlike previous population genetic models, which assume constant population sizes \citep{JordanConnallon2014,Olito2017}, mortality caused by inbreeding depression can strongly influence population persistence in our evolutionary demographic model. Populations with high fertility rates can maintain positive population growth rates despite this additional mortality. This causes a greater proportion of SA polymorphic parameter space to be demographically viable, with the demographic model predictions converging on those from population genetic models in high-fertility populations (fig.~\ref{fig:deltaPolySpace}).  \textcolor{blue}{Note, however, that in our model, mortality due to late-acting inbreeding depression also reduces the population effective selfing rate, and therefore shifts the invasion boundaries for the SA alleles to more closely resemble an outcrossing population. In some cases, this can lead to larger regions of viable polymorphic parameter space than predicted by the population genetic models, which can only include early-acting inbreeding depression (e.g., see cross symbols in fig.~\ref{fig:deltaPolySpace})}. However, as the selfing rate increases, demographic viability eventually crashes when the population can no longer sustain the concomitant increase in mortality due to inbreeding depression (fig.~\ref{fig:deltaPolySpace}). For populations with lower fertility rates, the proportion of demographically viable polymorphic parameter space \textcolor{blue}{is very similar to} the population genetic model predictions \textcolor{blue}{for low selfing rates}.

@@

The point in the life-cycle where inbreeding depression manifests can influence the threshold selfing rate at which demographically viable polymorphic parameter space crashes. Our results indicate that population viability was most sensitive to inbreeding depression affecting juvenile survival rates ($\delta_j$; fig.~\ref{fig:deltaPolySpace}, dark blue points), while early-acting inbreeding depression ($\delta$, ovule abortion shortly after fertilization) had a similar effect on population viability as late-acting inbreeding depression affecting adult survial ($\delta_a$) and juvenile-to-adult transition rates ($\delta_{\gamma}$). Inbreeding effects on juvenile survival had the strongest effect on population viability because on average individuals will spend multiple time steps in the juvenile stage before they mature. At each time step, juveniles have a probability $\gamma$ to mature and a probability $\sigma_j(1-\delta_j)$ to survive. Inbreeding depression at the juvenile stage therefore makes it harder to survive long enough to mature. \textcolor{blue}{Of course the strength of this effect is influenced by the maturation rate, $\gamma$, which therefore also affects the shape of the curves in Figure \ref{fig:deltaPolySpace}.} Although early-acting inbreeding depression ($\delta$) actually manifests earlier in the life-cycle than juvenile survival, it acts only once by influencing the total number of self-fertilized zygotes that become juveniles. Note, however, that the relative strength of inbreeding depression at different stages of the life history will depend on the interaction between the various inbreeding depression parameters, which we have precluded from our analyses.
